---
title: "Project"
author: "Lauren Cunild"
date: "12/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we need to download our data from the NYT data set. We are using the rolling average data for deaths and cases, and the mask use by county data set. Next, we need to join the data sets, so that we can better analyze whether cases and deaths do depend upon the mask use in the county, or state.

```{r}
# Import packages
library(tidyverse)
library(dplyr)
library(maps)
# Import data
mask_use <- read.csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/mask-use/mask-use-by-county.csv')
rolling_avg <- read.csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/rolling-averages/us-counties-2021.csv')
# Extract county id code
rolling_avg_new <- mutate(rolling_avg,geoid = as.integer(substr(geoid,5,9)))
# Mutate to rename county id code, so that the two data sets are consistent
rolling_avg_new <- rename(rolling_avg_new, COUNTYFP = geoid)
# create a new data set with both the cases/deaths averages and mask use data by county
data <- inner_join(rolling_avg_new, mask_use, by="COUNTYFP")
head(data)
```

We can now look at the mask frequency data by county, to identify any trends in our data set. We want to see which areas have higher and lower mask usage, and whether it depends on other factors, such as geographic location.

```{r}
# Extract mask use data
mask_data <- data[c("date", "COUNTYFP", "county", "state","NEVER","RARELY", "SOMETIMES", "FREQUENTLY", "ALWAYS")]
```

As the data set is very large, we have chosen three dates, approximately 6 months apart. We chose all three dates to be a Wednesday, as reporting can be delayed over the weekend, which would skew our data if some dates were weekends, and some were not.

```{r}
# filter for 3 different dates, approx 6 months apart
mask_data_filtered <- filter(mask_data, date == '2021-01-21'| date == '2021-06-09'| date == '2021-11-17')
mask_data_filtered
```
Next, we created a new variable called `mask_frequency`, which assigns a score to each county based upon the reported mask use. We gave never a score of 0.2, rarely 0.4, sometimes 0.6, frequently 0.8 and always 1. We then multiplied these scores by the probability, to give an approximate percentage of the amount of time people were masks for in the county, which higher scores corresponding to a higher reported mask usage.

```{r}
# create new df to assign mask usage score to each county
mask_freq_filtered <- mask_data_filtered %>% mutate(mask_frequency = 1*mask_data_filtered$ALWAYS + 0.8*mask_data_filtered$FREQUENTLY + 0.6*mask_data_filtered$SOMETIMES + 0.4*mask_data_filtered$RARELY + 0.2*mask_data_filtered$NEVER)
# display data in descending order
mask_freq_filtered <- arrange(mask_freq_filtered, desc(mask_frequency))
mask_freq_filtered <- select(mask_freq_filtered, c('date', 'COUNTYFP','county', 'state', 'mask_frequency'))
mask_freq_filtered
```
We also need to examine how the number of cases and deaths vary by county. To do this, we will use the number of cases and deaths reported per 100k of the population, to standardize across our data set. 

```{r}
# create df for cases and deaths avg per 100k in each county on chosen date
cases_data <- data[c("date", "COUNTYFP", "county", "state","cases_avg_per_100k", "deaths_avg_per_100k")]
cases_data_filtered <- filter(cases_data, date == '2021-01-21'| date == '2021-06-09'| date == '2021-11-17')
cases_data_filtered
```
Now that we have looked at the two data sets individually, we want to combine them, so that we can answer our question of whether mask use does affect the number of deaths and cases reported.

```{r}
# Join previous two df
cases_mask_data <- right_join(cases_data_filtered, mask_freq_filtered, by=c ("COUNTYFP", "date","county", "state"))
cases_mask_data
```
As there are over 3,000 counties, we will group the counties by state, to answer our hypothesis on a state-wide level.

Looking at the average number of cases by state:


```{r}
cases_mask_data %>%
  mutate(State = fct_lump(state))%>%
  group_by(State)%>% 
  summarise(case_avg = cases_avg_per_100k) %>%
  ggplot(aes(x=State, y=case_avg))+
  geom_point()+ theme(
        axis.text.x=element_blank())
```
For deaths per 100k, by state:

```{r}
cases_mask_data %>%
  mutate(State = fct_lump(state))%>%
  group_by(State)%>% 
  summarise(death_avg = deaths_avg_per_100k) %>%
  ggplot(aes(x=State, y=death_avg))+
  geom_point()+ theme(
        axis.text.x=element_blank())
```

For mask usage, by state:
```{r}
cases_mask_data %>%
  mutate(State = fct_lump(state))%>%
  group_by(State)%>% 
  summarise(mask_usage = mask_frequency) %>%
  ggplot(aes(x=State, y=mask_usage))+
  geom_point()+ theme(
        axis.text.x=element_blank())
```
We can see that although the average mask usage seems to be consistent across states, along with the highs and lows for each state, there is a large variation within the counties in each state. Therefore, we want to analyze the relationship between mask usage and deaths/cases on a county level. We can do this with a scatter graph:

```{r}
# draw stratified graphs with smooth line
cases_mask_data %>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=mask_frequency,y=cases_avg_per_100k))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)
```
It appears that we have a large outlier, with more than 1000 cases per 100k of the population. This is making it difficult to analyze any trend in our data, as the rest of our data is clustered at much lower average case numbers. We can look at this data point more closely:
```{r}
cases_mask_data %>%
  filter(cases_avg_per_100k>1000)
```
We can see that this point is for the county of Forest, Pennsylvania. Looking up this county in the case data set:
```{r}
data %>% filter(COUNTYFP == '42053' &  (date == '2021-01-19' |date == '2021-01-20' | date == '2021-01-21'| date == '2021-01-22'| date == '2021-01-23'))%>%
  select(c('date','county','cases'))
```
From looking at the reported number of cases just before and after the date with the large case average, we can see that it appears to be a reporting bias, rather than such a large jump in cases. Although it could be a large outbreak, we would expect there to be an increased number of cases in the days preceding and in the days after our chosen date, however this is not the case. Therefore, it looks like there could have been a delay in reporting, with cases from a longer period all being recorded on a single day, making the average much higher than it otherwise would have been. Therefore, we can exclude this data point from our analysis, to see if there is a clearer trend without our regression being skewed by the outlier. 

```{r}
# draw stratified graphs with smooth line
cases_mask_data %>%
  filter(cases_avg_per_100k<=1000)%>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=mask_frequency,y=cases_avg_per_100k))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)
```
Even after excluding our outlier, there still does not appear to be a clear trend between mask frequency and the average number of cases. However, we can see that mask usage is high across all counties, with only three below 0.5. Therefore, it is possible that in general people are wearing masks in high-risk situations, for example in crowded spaces on when indoors, and therefore wearing masks at other times, which would increase the frequency score does not cause a further decrease in case numbers. 

We can also look at the average cases and deaths in the counties which report the lowest and highest proportion of people who "never" wear a mask, and who "always" where a mask, to see if this gives a clearer answer to whether there is a statistical relationship between wearing a mask and getting Covid, or indeed dying from it. 

```{r}
full_cases_mask_data <- right_join(cases_data_filtered, mask_data_filtered, by=c ("COUNTYFP", "date","county", "state"))
full_cases_mask_data <- full_cases_mask_data %>% mutate(low_usage = full_cases_mask_data$NEVER + full_cases_mask_data$RARELY, high_usage = full_cases_mask_data$FREQUENTLY + full_cases_mask_data$ALWAYS)

# Plotting average cases per 100k against percentage who rarely or never wear masks
full_cases_mask_data %>%
  filter(cases_avg_per_100k<=1000)%>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=low_usage,y=cases_avg_per_100k))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)

# Plotting average cases per 100k against percentage who frequently or always wear masks
full_cases_mask_data %>%
  filter(cases_avg_per_100k<=1000)%>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=high_usage,y=cases_avg_per_100k))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)
```
Looking at the above scatter graphs, we can conclude that there does not seem to be a statistically significant correlation between mask usage and the average number of Covid cases.

Looking at deaths instead:
```{r}
# Plotting average deaths per 100k against percentage who rarely or never wear masks
full_cases_mask_data %>%
  filter(cases_avg_per_100k<=1000)%>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=low_usage,y=deaths_avg_per_100k))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)

# Plotting average deaths per 100k against percentage who frequently or always wear masks
full_cases_mask_data %>%
  filter(cases_avg_per_100k<=1000)%>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=high_usage,y=deaths_avg_per_100k))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)
```
Again, we do not seem to have a significant relationship between mask usage and the average number of deaths per 100k people. However, the counties with higher death rates do appear to report a lower mask usage. We can see that in the following graph:
```{r}
# Plotting higher average deaths per 100k against percentage who rarely or never wear masks
full_cases_mask_data %>%
  filter(deaths_avg_per_100k>5)%>%
  group_by(COUNTYFP)%>%
  ggplot(aes(x=low_usage,y=deaths_avg_per_100k))+
  geom_point()+geom_smooth(method="lm", se=FALSE)
```
Therefore, it seems like some other factors could be influencing death rate, causing it to remain low regardless of mask usage. We can look at death rate on a map, to see if geographic location (i.e. whether a county is rural or near a big city) affects death rate.



```{r}
county_df <- map_data("county")
full_cases_mask_data$subregion<-tolower(full_cases_mask_data$county)
map_mask_df<-right_join(county_df,full_cases_mask_data)

map_plot<-ggplot(map_mask_df)+ geom_polygon(aes(x=long,y=lat,fill= cases_avg_per_100k,group=group))+coord_fixed(1.3)+theme_bw()

remove_axis <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

map_plot +  scale_fill_gradientn(colours = rainbow(5) ,trans = "log10")+remove_axis
```

Seems like there is certain hotspots in the south and in Texas and California with high number of cases. Rural areas seem to have lower number of cases, but some urbans areas have low and others have high


```{r}
map_plot<-ggplot(map_mask_df)+ geom_polygon(aes(x=long,y=lat,fill= low_usage,group=group))+coord_fixed(1.3)+theme_bw()

map_plot +  scale_fill_gradientn(colours = rainbow(4), trans = "log10")+remove_axis



```

It seems like in rural areas people wear there mask less. The high the low_usage score the less they wear their mask.

```{r}
georgia_data<- map_mask_df%>%filter(region=="georgia")

map_plot<-ggplot(georgia_data)+ geom_polygon(aes(x=long,y=lat,fill= cases_avg_per_100k,group=group))+coord_fixed(1.3)+theme_bw()

map_plot +  scale_fill_gradient2()+remove_axis

```
There are a few areas in georgia which have a lot of cases but most seem to have few.

```{r}
state_df <- map_data("state")
full_cases_mask_data$region<-tolower(full_cases_mask_data$state)
map_mask_df<-inner_join(state_df,full_cases_mask_data,by=c("region"))


map_plot<-ggplot(map_mask_df)+ geom_polygon(aes(x=long,y=lat,fill=cases_avg_per_100k,group=group))+coord_fixed(1.3)+theme_bw()

map_plot + scale_fill_gradientn(colours = rainbow(4), trans = "log10")+remove_axis

```



